---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
    collapse = TRUE,
    comment = "#>",
    fig.path = "man/figures/README-",
    out.width = "100%",
    dpi = 300
)
```

# `{epl}` package

<!-- badges: start -->

<!-- badges: end -->

`{epl}` is a repository of functions and scripts for common tasks for use by students in the UBC Environmental Physiology Laboratory.

## Online App

`<under development>`

## Installation

You can install the development version of `{epl}` from [GitHub](https://github.com/jemarnold/epl) with:

``` r
# install.packages("remotes")
remotes::install_github("jemarnold/epl")
```

## Usage

#### `example_epl()`

This helper function can be used to retrieve example data files included in {epl} to test processing functions.

```{r}
#| error: TRUE
library(tidyr, warn.conflicts = FALSE)
library(dplyr, warn.conflicts = FALSE) ## for data wrangling
library(ggplot2) ## for plotting
library(epl) ## load {epl} package

## set the script default {ggplot2} display theme (see below for details)
theme_set(theme_epl())

## calling `example_epl()` will return a list of all included example files
example_epl()

## partial matching will error if matches multiple
example_epl("tymewear")

## calling a specific file by name will return the file path to that file
## partial string matching works for uniquely identifiable file names
example_epl("tymewear_live")
```

### `read_tymewear()`

This function will read *"Live Processed CSV"* or *"Post Processed CSV"* files exported from *Tymewear VitalPro* and return a list of two data frames with recorded data and file details.

-   `tyme$data` contains the recorded raw data samples for breathing rate, tidal volume, and ventilation: `c("br", "vt", "ve")`, with time in seconds and timestamps in `datetime` format.

-   `tyme$details` contains the file details & metadata, including `c("gender", "weight", "date", "app-start-time")`.

```{r}
## specify the file path via any number of methods
## this gives an example format
file_path <- file.path("root_project_folder", "data_folder", "my_file.csv")
file_path

## example files are available by calling `example_epl()`
file_path <- example_epl("tymewear_live")

tymelive <- read_tymewear(file_path)
tymelive

```

### `read_parvo()`

This function will read *.CSV* or *.xlsx* (but not *.XLS*) files exported by *Parvo Medics TrueOne 2400* and return a list of three data frames with recorded data, file details, and events.

-   `parvo$data` contains the recorded raw data samples for exported data channels (e.g. `c("VO2", "VCO2", "RR", "Vt")`) and calculated metabolic values (e.g. `c("O2kJ", "O2kcal", "Paer", "METS")`; see `?read_parvo` for details on calculated values).

-   `parvo$details` contains the file details & metadata, including `c("Date", "Name", "Sex", "Age")` (all metric values).

-   `parvo$events` contains any manually entered events with `"TIME"` in seconds and event descriptions.

```{r}
## example files are available by calling `example_epl()`
file_path <- example_epl("parvo_binned")

parvo <- read_parvo(file_path)
parvo
```

`read_parvo()` has an argument to add timestamps in `datetime` format to `parvo$data`, taken from `parvo$details` file start time. This is recorded from system time to the nearest whole second (i.e. precise to ± 0.5 sec, assuming system time is accurate).

```{r}
## timestamps can be added
read_parvo(file_path, add_timestamp = TRUE)$data
```

### `replace_outliers()`

This function can detect and replace local outliers in vector data with either `NA` or the local median value, within a window defined by `width`.

`width = 7` (sample window = `-7 < idx < 7`) seems suitable for breath-by-breath ventilatory data, however this can be modified depending on your data and to suit your desired response signal.

See `?replace_outliers` for more information.

```{r}

tyme_data <- tymelive$data
## filter breathing rate vector data for local outliers
vt_filtered <- replace_outliers(tyme_data$vt, width = 7, method = "median")

## plot raw signal and highlight outliers (see below for custom ggplot2 functions)
ggplot(tyme_data) +
    aes(x = time, y = vt) +
    ylab("Tidal Volume (L)") +
    scale_x_continuous(
        name = "Time (mm:ss)",
        breaks = breaks_timespan(),
        labels = format_hmmss
    ) +
    scale_colour_epl() +
    geom_line(aes(colour = "BR")) +
    geom_point(
        data = slice(tyme_data, which(vt_filtered != vt)),
        aes(y = vt, colour = "outliers"))

## to replace outliers from all channels in a data frame
tyme_clean <- tyme_data |> 
    mutate(
        across(c(br, vt, ve), \(.x) replace_outliers(.x, 7))
    )

## check relative (%) effect of removing outliers on signal SD
1 - sd(tyme_clean$vt) / sd(tyme_data$vt)

## and on sample noise
1 - mean(abs(diff(tyme_clean$vt))) / mean(abs(diff(tyme_data$vt)))

```

### `theme_epl()` & custom `{ggplot2}` plotting functions

This package includes a clean plot template and colour palette for `{ggplot2}`, the data visualisation engine of choice in R. We have already seen the basics of what it looks like.

The key function is `theme_epl()`, which can be called at the top of a script with `theme_set(theme_epl())` (as we have done at the top of this page) to set our plot defaults. Or it can be called within an individual plot, as below. See `?theme_epl` for details.

```{r}
parvo_data <- parvo$data

## explore what each setting does to the aesthetics of the plot
ggplot(parvo_data) +
    aes(x = TIME) +
    coord_cartesian(ylim = c(0, NA)) + ## set plot x- & y-axis limits
    theme(
        panel.grid.major.y = element_line() ## modify the current theme
    ) +
    scale_x_continuous( ## parameters for the x-axis
        name = "Time (mm:ss)",
        breaks = breaks_timespan(), ## pretty time x-axis formatting
        labels = format_hmmss ## pretty time x-axis formatting (see below)
    ) +
    scale_y_continuous( ## parameters for the y-axis
        name = "VO2  |  VCO2 (L/min)",
        expand = expansion(mult = 0) ## adjust space above/below max/min
    ) +
    scale_colour_manual( ## set the colour mapping
        name = NULL,
        aesthetics = c("colour", "fill"), ## for both colour & fill
        values = setNames(palette_epl()[c(6, 5)], c("VO2", "VCO2")), ## set custom colours (see below)
    ) +
    guides(
        colour = guide_legend(override.aes = list(alpha = 1)) ## modify the legend keys)
    ) +
    geom_area(aes(y = VO2, colour = "VO2", fill = "VO2"), 
              alpha = 0.1, key_glyph = "path") +
    geom_line(aes(y = VCO2, colour = "VCO2"))
```

`palette_epl()` includes some nice colours which can be called individually with `scale_colour_manual(values = palette_epl()[1:2])`, or as a default palette built into `scale_colour_epl()` (used in the two plots above). See `?scale_colour_epl` for more options.

```{r}
scales::show_col(palette_epl()) ## visualise the colour palette
```

Time values on the x-axis can be displayed in *"h:mm:ss"* units with `format_hmmss()`, and breaks set to pretty time-ish values (e.g. 60-second or 5-minute (300 sec) increments) with `breaks_timespan()`, as used in the plots above.

### `filter_data()`

`<under development>`

### `detect_peaks()`

`<under development>`

## To do

- Create vignette for main package usage.

- Update EPL Parvo App.

- Update `read_tymewear()` method for *"tymepost*" export file type.

- Add mean peak value detection, i.e. for V̇O~2~peak.

- Add digital filtering methods (e.g. Butterworth, smoothing spline, simple moving average, binning).

- Add 4-parameter monoexponential curve fitting via `nls()` self-starting functions.

- ~~Add template display theme for `ggplot2` plotting.~~

- ~~Add local outlier filtering for metabolic data.~~
